---
title: "Blocket Car Analysis"
author: 
  - Geisol Yissel Urbina and Maria Lagerholm
output: 
  github_document:
    toc: true
    toc_depth: 2
---


# Introduction

In this project, we explore the Swedish used car market using data scraped from Blocket and enriched with technical attributes from Transportstyrelsen and external statistics from SCB. The analysis has two core objectives:

To predict market prices for mid-range used cars in Sweden using machine learning (e.g., Random Forest).

To understand what variables actually influence insurance premiums and vehicle tax levels via statistical inference (e.g., linear regression).

We combine predictive and explanatory modeling to better understand price-setting mechanisms and cost structures for private car ownership. The figure below, generated from SCB data, illustrates trends in new car registrations, which reflect macro-level market shifts that may indirectly affect the availability and pricing of used cars.




---


```{r}

library(remotes)
#remotes::install_github("rOpenGov/pxweb", force = TRUE, build_vignettes = FALSE)
library(pxweb)

Sys.setlocale(locale = "UTF-8")

# Step 1: Read your JSON file into a proper query object
px_query <- pxweb_query("scb/trafik.json")

# Step 2: Download the data using the parsed query object
px_data <- pxweb_get(
  url = "https://api.scb.se/OV0104/v1/doris/sv/ssd/TK/TK1001/TK1001S/SnabbStatTK1001",
  query = px_query
)

# Step 3: Convert to data frame
px_data_frame <- as.data.frame(px_data, column.name.type = "text", variable.value.type = "text")


```
```{r}
library(ggplot2)
library(dplyr)

# Om du redan har laddat ner datan till t.ex. `px_data_frame`
# Annars kan du skapa ett exempeldata manuellt så här:
df <- px_data_frame

# Byt gärna till enklare kolumnnamn
colnames(df) <- c("manad", "forandring")

# Om du inte redan gjort detta:
df$datum <- as.Date(paste0(substr(df$manad, 1, 4), "-", substr(df$manad, 6, 7), "-01"))

ggplot(df, aes(x = datum, y = forandring)) +
  geom_point(size = 1.5) +
  geom_smooth(method = "loess", span = 0.2, color = "darkred", se = FALSE, size = 1) +
  theme_minimal() +
  labs(
    title = "Trend för förändring i nyregistrerade bilar",
    x = "Datum",
    y = "Förändring jämfört med året innan (%)"
  )



```

